---
title: "Data Science R Assessment"
author: "Yves Greatti"
date: "2023-08-19"
output:
  html_document:
    theme: united
    code_folding: hide
    highlight: tango
    df_print: tibble
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Data Characteristics

The dataset encompasses crucial economic information for each country
and continent worldwide, spanning five-year intervals from 1962 to 2007.
This data pertains to:

1.  GDP indicators (such as GDP per capita, exports, imports,
    contributions from various sectors like industry, agriculture,
    services, and the financial sector, as well as inflation rates).
2.  Energy metrics (including electric power consumption and overall
    energy usage).
3.  CO2 emissions data.
4.  Societal metrics like life expectancy and fertility rates.
5.  Population statistics, including total population and population
    density.


## CO2 Emissions vs. GDP per Capita

```{r echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls())
library(dplyr)
library(ggplot2)
library(plotly)
library(kableExtra)

setwd("~/github/skill-assessments/R for Data Science")
df <- as_tibble(read.csv("gapminder_clean.csv"))
df <- df %>% select(-1)

head(df) %>%
  kbl() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

The correlation between CO2 emissions (metric tons per capita) and GDP
is highly significant, demonstrating a robust statistical strength, as
evidenced by a correlation coefficient of 0.93 and an almost negligible
p-value.

```{r message=FALSE,  warning=FALSE, echo=FALSE}
df2 <- df %>% 
   filter(Year == 1962) %>%
  filter(!is.na(CO2.emissions..metric.tons.per.capita.)) %>%
  mutate(continent = factor(continent)) 

# Calculate correlation coefficient and p-value
pear_res <- cor.test(df2$gdpPercap, 
                df2$CO2.emissions..metric.tons.per.capita.,
                method = "pearson")

print(paste0("Correlation between CO2 emissions and GDP is:", round(pear_res$estimate,3), 
             " with a p-value:", pear_res$p.value))

spear_res <- cor.test(df2$gdpPercap, 
                     df2$CO2.emissions..metric.tons.per.capita.,
                     method = "spearman")
```

```{r message=FALSE, echo=FALSE, warning=FALSE}
p <- df2 %>%  ggplot(aes(x = gdpPercap, y = CO2.emissions..metric.tons.per.capita.)) +
  geom_point(aes(size = CO2.emissions..metric.tons.per.capita., color = continent)) +
  geom_text(aes(label = Country.Name), nudge_x = 300, nudge_y = 0.01, size = 2) +  
  scale_x_continuous(limits = c(0, 25000)) + 
  scale_y_log10(labels = scales::label_number(scale = 1e3, suffix = "K")) +  
  labs(x = "GDP per Capita", y = "CO2 Emissions (metric tons per capita), LOG10",
       title = "CO2 Emissions vs. GDP per Capita",
       size = "CO2 Emissions",
       color = "Continent") + 
  scale_color_manual(values = c("Asia" = "blue", "Europe" = "green", "Africa" = "red",
                                "Americas" = "purple", "Oceania" = "orange")) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12)
  ) +
  geom_smooth(method = "lm", se = F, size=.5, alpha=0.5) +
  geom_text(aes(x=15000, y=0.05, 
                label = paste0('Pearson:', 
                               round(pear_res$estimate,2),
                               ', P-value < 0.001\n',
                               'Spearman:', 
                               round(spear_res$estimate,2),
                               ', P-value < 0.001')
                ), 
                color = 'blue',  size=3)
p
```

```{r message=FALSE, echo=FALSE, warning=FALSE}
df3 <- df %>%
  group_by(Year) %>%
  summarize(correlation = cor(CO2.emissions..metric.tons.per.capita., 
                              gdpPercap, 
                              use = "complete.obs")) %>%
  filter(correlation == max(correlation)) 
cat("The year with the strongest correlation between CO2 Emmissions
    and GDP Per Capita is:", df3$Year)

```

**Interactive plot**

```{r message=FALSE, echo=FALSE, warning=FALSE}
p2 <- p +
  guides(color = FALSE, size = FALSE)
ggplotly(p2)
```
 ## Relationship Between Continent and Energy use (kg of oil equivalent per capita)
 
To understand the relationship between the categorical variable
("continent") and the continuous variable ("Energy use (kg of oil
equivalent per capita)"), we can use analysis of variance (ANOVA) or
perform a visual analysis using violin plots.

```{r message=FALSE, echo=FALSE, warning=FALSE}
model <- lm(Energy.use..kg.of.oil.equivalent.per.capita. ~ continent, data = df)
anova_result <- anova(model)
cat(anova_result$`Pr(>F)`[1])
```

This extremely low p-value indicates that there is a statistically
significant relationship between"continent" and "Energy use (kg of oil
equivalent per capita)."

```{r message=FALSE, echo=FALSE, warning=FALSE}
pl <- df %>% ggplot(aes(x = continent, y = Energy.use..kg.of.oil.equivalent.per.capita.)) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  labs(x = "Continent", y = "Energy use (kg of oil equivalent per capita)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplotly(pl)
```

## Comparing Europe and Asia's 'Imports of goods and services (% of GDP)' after 1990
\
Since there's only one measurement available for each country per year,
the total importation used in Europe and Asia remains independent. Given
this independence, we can choose a paired t-test to analyze any
potential differences between the energy usage in the two regions.

```{r message=FALSE, echo=FALSE, warning=FALSE}

total_import_per_country <-  df %>%
  filter(Year > 1990) %>%
  group_by(Year, continent, Country.Name) %>%
  filter(continent != "") %>%
  summarize(SumImport = sum(Imports.of.goods.and.services....of.GDP., na.rm = TRUE)) %>%
  filter(Country.Name != "" & SumImport != 0) %>%
  distinct(Year, Country.Name, continent, .keep_all = TRUE)  %>%
  ungroup()
  
# Group by Year and continent, then summarize to compute total imports per continent
total_import_per_continent <-  total_import_per_country %>%
  group_by(Year, continent) %>%
  summarize(SumImport = sum(SumImport, na.rm = TRUE))

# Filter data for each continent
europe_data <- total_import_per_continent %>%
  filter(continent == "Europe") %>%
  pull(SumImport)

asia_data <- total_import_per_continent %>%
  filter(continent == "Asia") %>%
  pull(SumImport)

tres <- t.test(europe_data, asia_data)
tres

#Create a bar plot using ggplot2
pl <- ggplot(total_import_per_continent %>% filter(continent %in% c("Asia", "Europe")),
             aes(x = Year, y = SumImport, fill = continent)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Total Importations", title = "Total Imports By Asia and Europe") +
  scale_fill_manual(values = c("Asia" = "blue", "Europe" = "green")) + 
  scale_x_continuous(breaks = seq(1992, 2007, by = 5)) +
  theme_classic()

ggplotly(pl)

```

Due to the high p-value obtained from the t-test, we can
conclude that there no substantial and statistically significant
distinctions in the total importations of goods and services between Europe and
Asia, spanning the time frame from 1990 onward.

## Identifying The Country with The Largest 'Population Density' Across All Years

```{r message=FALSE, echo=FALSE, warning=FALSE}

highest_density_by_year <- df %>%
  group_by(Year, Country.Name) %>%
  summarize(max_pop_density = max(Population.density..people.per.sq..km.of.land.area., na.rm = TRUE)) %>%
  arrange(Year, desc(max_pop_density)) %>%
  group_by(Year) %>%
  slice(1)


plt <- ggplot(highest_density_by_year, aes(x = Year, y = max_pop_density, color = Country.Name)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Max Population Density", 
       title = "Highest Population Density by Year",
       color='Country') +
  theme_minimal() +
  #theme(legend.position = "none")
  theme(legend.position = "bottom") +
  geom_text(aes(label = Country.Name), nudge_y = 5, show.legend = FALSE)   +
  scale_x_continuous(breaks = seq(1967, 2007, by = 5)) 
ggplotly(plt)
```

## Country With the Greatest Increase in 'Life expectancy at birth, total (years)' between 1962 and 2007?
  